Title: Fish do not need Visas: the transboundary nature of world fished species

### 2019-11-20 Manuscript_Draft.Rmd
Updates to manuscript

---
title: "Title: Fish do not need Visas: the transboundary nature of world fished species (70/96 character maximum)"
authors: "Authors: Juliano Palacios-Abrantes^1^, Gabriel Reygondeau^1,2^, William W.L. Cheung^1^"
date: "Affiliations: ^1^ Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada, ^2^ Yale"
# subtitle: "Fish do not need Visas (16/40 character maximum)"
output:
  pdf_document:
    fig_caption: yes
  word_document:
    reference_docx: ./Manuscript_Instructions/science_template.docx
geometry: margin=1in
editor_options:
  chunk_output_type: console
bibliography: ./References/FishForVisa_Ref.bib
csl: ./References/science.csl
---

@@

<!-- # Introduction (~600 words) -->

Species are not randomly distributed in the world. The area in space and time where a stable population lives and reproduces is determined by the species' fitness response to a series of biotic and abiotic factors [@Hunch] and their influence in the evolutionary process [@REF]. Species distributions are key elements in marine resources management as human-made spatial boundaries are a major part of the management system [@Song:2017iua]. However, incorporating the concept of species distribution has not allays been the case in fisheries management, nor policy [@Alexa]. The delimitation of Economic Exclusive Zones (EEZs) in the arly 80's virtually splitted the distribution of many species, creating shared species between nations, and adding another level of complexity to fisheries management. However, 40 years after their establishment, the question of how many shared species exists in the world based on their distribution is still unanswered. In here, we take the most detailed database of marine fished species to show that many species are shared, representing an important contribution to national economies and food security, and highlighting the importance to acknowledge this in the policy arena.

From 1973 to 1982 the members of the United Nations held a series of meetings to discuss regulations regarding the high seas, a region of international common property, which, at that time, represented waters after 12 miles from shore [@McRae and Munro, 1989]. The establishment of the UN Convention on the Law of the Sea (UNCLOS) allowed coastal states to claim jurisdiction over 200 nautical miles of their coasts [@UN:1982]. Despite intended to improve fisheries management by granting property rights over shared resources, this arbitrary delimitation of management area had no input from biogeography; ignoring species distributions created what we know today as shared stocks [@REF]. The Food and Agricultural Organization (FAO) recognizes three types of shared stocks: (*i*) transboundary, those stocks that are shared by 2 or more neighboring coastal nations; (*ii*) straddling, those stocks that are shared by two or more nations and also the high seas; and (*iii*) highly migratory stocks; species that are found in the EEZs of coastal nations that are not necessarily neighbors, and the high seas.

The origination of shared stocks required the establishment of new fisheries management methods. Managers and scientists adopted a game theory approach where collaboration would most likely result in the best overall outcome for nations sharing a common resource [@REF]. Lack of collaboration in shared stocks threatens the stock's sustainability [@REF], reduces the profitability potential of the fishery [@REF], and creates conflict between coastal nations [@REF-Jessica?]. Moreover, new emerging conflicts are expected as climate change shifts the distribution of marine species and new shared fisheries emerge [@Pinsky:2018cb].

There is a gap in assessing transboundary species in the world based on their spatial distribution. It was estimated in 1997 that there could be up to 1,500 transboundary fish species in the world, however, such estimation lacked a proper assessment due to limited information at the time [@Caddy:1997ue]. More recently, it was estimated there were 344 shared species, representing a capture of 34.2 × 10^6^ t and a global landed value of USD 30.7 × 10^9^ [@Teh:2015gd]. While this represents a good baseline to start, the analysis was based on a literature review of shared species and did not include a mechanistic way of determine the transboundary nature of the species [@Teh:2015gd]. Therefore, 40 years after UNCLOS, the question of how many transboundary species exists in the world is still unanswered. This is of uppermost need as the fisheries management approach differs substantially between EEZ-discreet and shared species [@Mills].

 <!-- **Results and discussion** (1000 words) -->

We combined a series of datasets and species distribution models to estimate the number of transboundary fished species and their contribution to food security and economics of the world. We adopted the UN definition of transboundary stocks (shared by neighboring EEZs) but carried the analysis at the species level, rather than stocks (populations within a species), due to lack of stock-specific spatial and ecological information on all of the fished species. Our results suggest that there are `r Total_trans_n` transboundary species in the world, wich is **something** related to what Caddy (1997) estimated and more than double of that of Teh and Sumaila (2010) [@Teh].

Tranboundary fisheries are quite important worldwide in terms of both landings and revenue as they are responsable for XXX tones of landings generating USD `r Total_Trans_Rev` million in fishing revenue. This represents XX% of the catch and `r round(Per_Trans_Rev,2)`% of total revenue generated by those fisheries in all EEZs of the world. All these values are higher than previouslly reported by Teh and Sumaila (2010) up to 20006 where the maximum amount was 45% in the mid-1990s [@Teh]. The importance of these species exceeds in EEZs like the United States, Peru and Norway where fewer species are responsable for higer landing values (Fig 1).

@@

Transboundary fisheries are vast in regions whith historic fisheries-related conflicts [@Spikjers]. Regional variation show higher frequency of transboundary species within EEZs along the South China sea, western central Atlantic, and the Mediterranean sea, followed by the Eastern north Atlantic (Fig 1). Some of these regions have been identified as most conflictive since 1970s with conflicts arrousing to even armed conflicts. Moreover, as climate change shifts species distributions and new transboundary species arrose [@Pinsky] as seen in past cases like the Pacific salmon [@REF], or the Macrrell waars [@Spijkers]. The joint management of shared stocks could be key to not only sustain fisheries into the future but also avoid international conflict [@REF].

Management of shared stocks have to be something. Although catch trends do not reflect stock status [@REf_Haters], they can inform provide a panoram of the situation when no other data is available [@Pauly]. In the case of transboundary species, we did not find a strong pattern suggesting a specific case (Fig 2).

@@

All species that share over >100 countries are widely distributed pelagic species such as Wahoo (*Acanthocybium solandri*), Common thresher shark (*Alopias vulpinus*), and tunas (*Thunnus sp.*) with a median of 65, 42, and 38 species per pelagic oceanic, bathymetric and nerictic species, respectively. The median for all other ecosystem preferences is under 35 as many of these species are not so widely distribute globally.

@@

@@

In here we present a novel method to estimate whether or not a species is transboundary. Our method is based on a series of criteria to be met and the overlap of species distributions and the spatial boundary of the world's EEZs. For the current analysis we work at the species level and adopt FAO's definition of transboundary species as those that occur within the Exclusive Economic Zone (EEZ) of two or more neighboring countries [@Munro:2002uf; @FAO]. All data was scaled at a 0.5 degree latitude x 0.5 degree longitude grid cells (*n* = ~180,000 grid cells).

@@

To determine the pool of transboundary marine species exploited by fisheries within each of the world's EEZs we first extract all the commercial marine fish or invertebrate species from the *Sea Around Us* database (http://www.seaaroundus.org and determined their current distributions. For this, we used four species-distributions data sources based on (*i*) observational data, (*ii*) an Environmental Niche Model (ENMs), (*iii*) a life-history-based distribution model, and (*iv*) fisheries catch data (**Table x**). Each source represent a different method of estimating the distribution of a given species and thus, provides a more robust result. Only those commercial fished species that had data from all four sources were included in the analysis, and therefore the final dataset comprised XXX species, XX% of the reported marine species in the SAU database and XX% of the reported taxa of the FAO (**See S1_Data**).

@@

*ii* The ENM-Nereus consists in a multimodel approach based on a Bioclim and a Bososted Regression Tree model [@Thuiller:2009gp], a Maxtent model [@Phillips:2006ff], and a Non-Parametric Probabilistic Ecological Niche Model [@Beaugrand:2011fd]. Environmental variables modeled were sea surface temperature, surface pH, surface oxygen concentration, and vertically integrated (0–100 m) net primary production (NPP) [@Asch:2018ca]. The ENM-Nereus employed all of the observational data-sources previously mention. Global environmental conditions were obtained and results averaged from three Earth System Models developed by the Geophysical Fluid Dynamics Laboratory's (GFDL- https://www.gfdl.noaa.gov/earth-system-model/), the Institute Pierre Simon Laplace (IPSL- www.icmc.ipsl.fr/), and the Max Planck Institute for Meteorology (MPI- www.mpimet.mpg.de/en/science/models/). For each species, we only considered models with an area under the curve larger than 0.7 and outputs from the models were average between *1970* and *2000* [@Asch:2018ca; @Reygondeau:2019uh].

*iii* The SDM-SAU model follows a five-steps process based on species-specific life history information, rather than environmental variables [@Close:2006ux; @Palomares:wg]. For each commercial fish species, the model first uses the FAO major fishing areas and countries EEZs to determine a broad distribution. It then uses life history information to delimit its range within the FAO fishing area (e.g. thermal preference, depth limits). The range is delimited even further by expert-review polygons and compared with that of AquaMaps [@Kaschner:2016tl], OBIS and GBIF occurrence data. The model then determines a species habitat preference based on the assumptions that the relative abundance of a species is determine by the number of habitats in a grid cell and the distance of the species to each habitat, and the importance of the habitat to the species size. Finally, the species equatorial submergence (e.g. the latitudinal region where a species is not seen in between poles) is estimated for each species.

@@

data.frame(Dasets_names,
           Brieff_Description,
           Sources) %>% 
  kable("latex", booktabs = T,
        caption = "Summary table of data sources ussed in the current analysis") %>%
  kable_styling(full_width = F) %>%
  column_spec(2,width = "8cm") %>% #wraps text in column2
  column_spec(3,width = "3cm") #%>% #wraps text in column2

@@

### Criteria 1; Neighbouring EEZs

As previously mentioned, we define transboundary species as those marine species that happen within the EEZs of two or more neighboring countries. Hence, the first criteria was that the species would only be transboundary between two neighboring countries, regardless of the species extended distribution. This way, a species such as Atlantic cod (*Gadus morhua*) distributed along the north Atlantic, was only considered transboundary between each of the neighboring nations covering its distribution, rather than the region as a whole. We assume that this would reduce the differences between species and stock as is more likely that neighboring species belong to the same stock, rather than, lets say, Cod in Norway and Canada. However, the analysis here presented was kept at the species level.

We define the EEZ boundaries using the SAU shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) that sub-divides EEZs by regions (e.g. Mexico Pacific and Mexico Atlantic) and determine the intersections between polygons using the *sf* *R* package [@SF:package]. When estimating transboundary species, we filter out those shared by the EEZs sub-regions, and when aggregating by country we only counted the species once, if it appears in more than one sub-region. Species that were present in EEZs that were non-continental territories neighboring other countries were kept (e.g. Argentina and Falkland Islands), but removed when the non-continental territory belonged to the same nation (e.g. Brazil and Fernando de Noronha).

@@

We assume that a species was only present in a given grid-cell if it was reported in the SAU catch database. Therefore, all species that were not reported as caught in any single year between the reference years (1970 to 2000) in a given grid-cell were dropped, regardless of the *Species Index*. The assumption relies in that if a commercial species is projected within any fishing country, such species would have been fished, and thus reported at some period of time, and thus validating the models.

@@

Fisheries are important contributors to the world economy and livelihoods. We estimated the economic contribution of transboundary species for each country using global ex-vessel price [@Tai2017]. The data includes ex-vessel price from multiple sources and a structured interpolation method (e.g. similar countries, species) to fill in data gaps [@Sumaila:2015uc]. Additionally, we used a previously estimated country-level multiplyer to understand the total direct, indirect, and induced impact of transboundary species on each country [@Dyck:2010ez]. A continent average was used when no country-level multiplyer was available.

We use the SAU catch data to estimate catch trends over time and determine the potential exploitation status of each species within a EEZ [@Grainger:1996tk; @Kleisner:2011wn]. Despite a possible overestimation of overexploited and collapsed stocks when using only catch data [@Branch:2011br], this method can still shed a light on stock status when only catch data is available [@FAO:2011vz; @FAO:2018KK; @Kleisner:2011wn]. However, the results should be seen as potential species status, rather than the true status of each species. We only assessed species within each EEZ that had at least ten years of data between the first and last reported landings with at least five consecutive data years, and their catch was more than 1,000 tonnes. Five categories were drown from the method depending on the year's catch, the maximum catch over time and the minimum catch after the catch peak (**Table XX**). Finally, we reported the predominant status over the last ten years of catch data (2004-2014).

@@

Stock_Status <- c("Rebuilding","Developing","Max Exploited","Over Exploited","Collapsed","No Status")

Rules <- c(
  "(Year of Catch > Year Post Max. Min. & Post Max Min Catch < (Max Catch*0.10)) & (Catch > (Max Catch*0.10) & Catch < (Max Catch*0.50))", # Reb.
  "Year of Catch < Year of Max. Catch & Catch <= (Max Catch*0.50) or Year of Max Catch = Last Year of data)", #Dev.
  "Catch > (Max Catch*0.50)", #Max Exp.
  "Year of Catch > Year of Max. Catch & (Catch > (Max. Catch*0.10) & Catch < (Max. Catch*0.50)", # Over.
  "Year or Catch > Year of Max. Catch & Catch < Max. Catch*0.10", # Colla.
  "Non of the above rules applyed"
)

>>>>

---
title: "Title: Fish do not need Visas: the transboundary nature of world fished species"
authors: "Authors: Juliano Palacios-Abrantes^1^, Gabriel Reygondeau^1,2^, William W.L. Cheung^1^"
date: "Affiliations: ^1^ Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada, ^2^ Yale"
# subtitle: "Fish do not need Visas (16/40 character maximum)"
output:
  word_document:
    reference_docx: ./Manuscript_Instructions/science_template.docx
  pdf_document:
    fig_caption: yes
geometry: margin=1in
editor_options:
  chunk_output_type: console
bibliography: ./References/FishForVisa_Ref.bib
csl: ./References/science.csl
---

@@

<!-- # Introduction (~555/500 words) -->

Species are not randomly distributed in the world. The area in space and time where a stable population lives and reproduces is determined by the species' fitness response to a series of biotic and abiotic factors [@Hunch] and their influence in the evolutionary process [@REF]. Species distributions are key elements in marine resources management as human-made spatial boundaries are a major part of the management system [@Song:2017iua]. However, incorporating the concept of species distribution has not allays been the case in fisheries management, nor policy [@Alexa]. The delimitation of Economic Exclusive Zones (EEZs) in the early 80's virtually spitted the distribution of many species, creating shared species between nations, and adding another level of complexity to fisheries management. However, 40 years after their establishment, the question of how many shared species exists in the world based on their distribution is still unanswered.

From 1973 to 1982 the members of the United Nations held a series of meetings to discuss regulations regarding the high seas, a region of international common property, which, at that time, represented waters after 12 miles from shore [@McRae and Munro, 1989]. The establishment of the UN Convention on the Law of the Sea (UNCLOS) allowed coastal states to claim jurisdiction over 200 nautical miles of their coasts [@UN:1982]. Despite intended to improve fisheries management by granting property rights over shared resources, this arbitrary delimitation of management areas had no input from bio-geography; ignoring species distributions created what we know today as shared stocks. The Food and Agricultural Organization (FAO) recognizes three types of shared stocks: (*i*) transboundary, those stocks that are shared by 2 or more neighboring coastal nations; (*ii*) straddling, those stocks that are shared by two or more nations and also the high seas; and (*iii*) highly migratory stocks; species that are found in the EEZs of coastal nations that are not necessarily neighbors, and the high seas. The origination of shared stocks required the establishment of new fisheries management methods. Managers and scientists adopted a game theory approach where collaboration would most likely result in the best overall outcome for nations sharing a common resource [@Bailey:2010ga]. Lack of collaboration in shared stocks threatens the stock's sustainability [@REF], reduces the profitability potential of the fishery [@REF], and creates conflict between coastal nations [@REF-Jessica?].

There is a gap in assessing transboundary species in the world based on their spatial distribution. It was estimated in 1997 that there could be up to 1,500 transboundary fish species in the world, however, such estimation lacked a proper assessment due to limited information at the time [@Caddy:1997ue]. More recently, it was estimated there were 344 shared species, representing a capture of 34.2 × 10^6^ t and a global landed value of USD 30.7 × 10^9^ [@Teh:2015gd]. While this represents a good baseline to start, the analysis was based on a literature review of shared species and did not include a mechanistic way of determine the transboundary nature of the species [@Teh:2015gd]. And so, despite recent research highlighting the interconnection of marine species [@Nandini], transboundary species are yet to be identified at a global scale. This is of uppermost need as the fisheries management approach differs substantially between EEZ-discreet and shared species [@Mills]. Moreover, international conflicts over shared resources are expected as climate change shifts the distribution of marine species and new shared fisheries emerge [@Pinsky:2018cb].

 <!-- **Methods summary** (250/250 words) -->

In here, we overlap the known distribution of **968** marine fish species with **256** Economic Exclusive Zones of the world to map the number of transboundary species and and their contribution to global catch and landed value. We adopted the UN definition of transboundary stocks (shared by neighboring EEZs) but carried the analysis at the species level, rather than stocks (populations within a species), due to lack of stock-specific spatial and ecological information on all of the fished species [@TEH].

We combined multiple species distribution data-sources and developed a series of criteria to determine if a species was transboundary. Current species distributions were obtained from an observational database of world marine species [@REFNEreus] and two structurally different species distribution models [@REFGabs; @Zeler]. The first criteria to determine if a species is transboundary was related to the agreement of the data-sources regarding the presence of a species in a single grid-cell. The second criteria was established to confirm the presence of a species by using a specially detailed catch data [@Zeler]. Finally, the third criteria accounted for the spatial distribution of the species between the neighboring EEZs. In addition to present the number of transboundary stocks, we estimated their contribution in terms of fishing revenue and landings and categorize them depending on their catch trend [@Pauly 2012]. Our study provides a first empirical approximation to understand the true transboundary nature of the world fisheries, a world that is deeply connected by larval dynamics [@nandini], fishing fleets [@Kurosawa_Juanito], and shared management [@RFMOS].

<!-- **Results and discussion** (~500/1000 words) -->

As Caddy (1997) pointed out, virtually all coastal nations share at least one transboundary species with their neighbors with some sharing over 100 species with surrounding EEZs (Fig. 1). With the current state of knowledge of fished species and their distributions, we estimate a total of `r Total_trans_n` transboundary species in the world, more than double of what had been previously estimated [@Teh:2015gd]. These fisheries are important in terms of both landings and revenue. In the last ten years, the global fleet has landed an average of XXX tones, generating USD `r Total_Trans_Rev` million in fishing revenue from shared stocks fished within EEZs. This represents **XX**% of the global catch and `r round(Per_Trans_Rev,2)`% of total revenue generated by those fisheries in all EEZs of the world. All these values are higher than previously estimated [@Teh]. The importance of these species exceeds in EEZs like the United States where **XX** species are responsible for up to **XX** millions USD and Peru where **17** species are responsible for XXX million USD (Fig 1). Transboundary fisheries are vast and economically important in regions with historic fisheries-related conflicts. The Asian region has seen 43.0% of all international fisheries conflicts since 2000; Europe, where 11.5% of fishing conflicts occur since 1974s [@Spijkers:2019cz], has been involved in a conflict between Ireland and the European Union over mackerel since 2007 [@Spijkers:2017ij], and finally, the Mediterranean has seen conflicts since the 1980's [@Kliot:1989kv]. The identification of current transboundary species is a first step towards establishing joint management plan of marine resources [@Link:2010ei], key to not only sustain fisheries but also avoid international conflict.

@@

![**Number of transboundary species by EEZ and their production by country**](./Figures/Fig1.png)

The two regions of the world with most shared species (North East and Central Atlantic) present quite different management scenarios. For the North East Atlantic, fisheries are managed by the European Union and quotas allocated by species and regulatory area, with some of them (e.g. Iceland and Norway) potentially having good fisheries management [@Melnychuk:2017gh]. In the other hand, fisheries along the west coast of Africa need to be improved, with countries like Nigeria where stock status, management, and enforcement is lacking [@Melnychuk:2017gh]. Ironically, this region has long been exploited by foreign European fleets with questionable fishing rights systems [@Mills:2004dj] and even some illegal activities [@Belhabib:2015bl]. Thus, just having joint management plans for shared stocks between neighbors will not be enough to achieve the sustainable goals of SDGs [@]. Countries need to work together towards a more equitable system of global fishing rights, combat illegal fishing both within their EEZs and outside them, and address global threats such as climate change and overfishing.

@@

Empirical analysis suggest that cooperation will have better outcomes for shared fisheries management. Examples vary between regions and species including Mexico and the United States over Pacific anchovies [@CisnerosMontemayor:2020kv], Norway and Russia over Atlantic cod (*Gadus morhua*)[@Diekert:2010gp], and between Namibia and South Africa over hake (*Merluccius spp*) [@Sumaila:2003vw]. However, few evidence exists today regarding the true efficiency of transboundary fisheries plans. Indeed, there are some successful bilateral management plans for species such as Pacific Halibut between Canada and the US as well [@REference?] and multilateral treaties like the Pacific Nauru Agreement that manages 30% of global skipjack tuna in the South Pacific Islands [@Aqorau:2018bh]. However, the effectiveness of Regional Fisheries Management Organizations, responsible for managing high migratory stocks in areas beyond national jurisdiction, has been previously questioned [@CullisSuzuki:2010fi]. Countries that are today harvesting species at rates less than 50% of the historical maximum catch (Category C) should be paying extra care in these species (Fig. 2). While these catch trends do not reflect stock status [@@Branch:2011br], they could reassemble fisheries that are over exploited or collapsed [@Kleisner:2011wn]. The rebuilding of a transboundary species could be strongly impacted by uncoordinated efforts of countries harvesting the same species but with different objectives [@Ref].

Management of transboundary fisheries needs to be looking forward. As species shift their known distributions to cope with a changing ocean, new transboundary species will arose, challenging the management of shared stocks worldwide [@Pinsky]. Conflicts in regions with high abundance of transboundary species will likely be exacerbated by climate change, as already seen in the dispute of Ireland and the European Union over Mackerel (*species*) [@Spijkers]. Understanding how climate change will affect the biogeography of marine species is a key element for proper spatial-based fisheries management, not only at the EEZ level, but also other levels of spatial tools [@Alexa, @Song:2017iua].

Species can have a cosmopolitan distribution, that is, a broader distribution, or be limited to a specific region, known as endemic [@REF]. Maybe not surprisingly, highly migratory pelagic-oceanic species such as Wahoo (*Acanthocybium solandri*), Common thresher shark (*Alopias vulpinus*), and tunas (*Thunnus sp.*) are among the more shared resources in the world with a median of 40 nations sharing these resources. These species are very important in terms of economic value as they (global) average revenue exceeds USDXXX. The median for species of all other ecosystem preferences is close to, or less than, 20, as many of these might be cosmopolitan but not necessarily highly migratory (e.g. *Solea solea*, *Panulirus argus*) (Fig. 3). However, less distributed species must not be ignored. The world of fisheries is highly connected trough larvae dispersal [@Ramesh:2019va] and so, even if a population swims between the borders of two nations, their larvae could be supplying populations thousands of kilometers away [@Ramesh:2019va]. Moreover, many coastal communities, specially in developing regions, and highly dependent on fishing local fish (e.g. reef fish) for food security and improved livelihoods [@CisnerosMontemayor:2016gq; Wabnitz:2018gf].

@@

![**Number of countries shared by transboundary species. Showing only species that share > 20 countries**](./Figures/Fig3.png)

<!-- Overview of West African fisheries under climate change: Impacts, vulnerabilities and adaptive responses of the artisanal and industrial sectors -->
<!-- https://www.sciencedirect.com/science/article/pii/S0308597X16302901#bib4 -->

@@

In here we present a novel method to estimate whether or not a species is transboundary. Our method is based on a series of criteria to be met and the overlap of species distributions and the spatial boundary of the world's EEZs. For the current analysis we work at the species level and adopt the Food and Agricultural Organization (FAO) definition of transboundary species as those that occur within the Exclusive Economic Zone (EEZ) of neighboring countries [@Munro:2002uf; @FAO]. All data was scaled at a 0.5 degree latitude x 0.5 degree longitude grid cells (*n* = ~180,000 grid cells).

@@

To determine the pool of transboundary marine species exploited by fisheries within each of the world's EEZs we first extract all the commercial marine fish or invertebrate species from the *Sea Around Us* database (http://www.seaaroundus.org and determined their current distributions. For this, we used four species-distributions data sources based on (*i*) observational data, (*ii*) an Environmental Niche Model (ENMs), (*iii*) a life-history-based distribution model, and (*iv*) fisheries catch data (**Table x**). Each source represent a different method of estimating the distribution of a given species and thus, provides a more robust result. Only those commercial fished species that had data from all four sources were included in the analysis, and therefore the final dataset comprised XXX species, 100% of the reported marine species in the SAU database and XX% of the reported taxa of the FAO (**See S1_Data**).

@@

*ii* The ENM-Nereus consists in a multimodel approach based on a Bioclim and a Bososted Regression Tree model [@Thuiller:2009gp], a Maxtent model [@Phillips:2006ff], and a Non-Parametric Probabilistic Ecological Niche Model [@Beaugrand:2011fd]. Environmental variables modeled were sea surface temperature, surface pH, surface oxygen concentration, and vertically integrated (0–100 m) net primary production (NPP) [@Asch:2018ca]. The ENM-Nereus employed all of the observational data-sources previously mention. Global environmental conditions were obtained and results averaged from three Earth System Models developed by the Geophysical Fluid Dynamics Laboratory's (GFDL- https://www.gfdl.noaa.gov/earth-system-model/), the Institute Pierre Simon Laplace (IPSL- www.icmc.ipsl.fr/), and the Max Planck Institute for Meteorology (MPI- www.mpimet.mpg.de/en/science/models/). See [@Asch:2018ca; @Reygondeau:2019uh] for model's details.

*iii* The SDM-SAU model follows a five-steps process based on species-specific life history information, rather than environmental variables [@Close:2006ux; @Palomares:wg]. For each commercial fish species, the model first uses the FAO major fishing areas and countries EEZs to determine a broad distribution. It then uses life history information to delimit its range within the FAO fishing area (e.g. thermal preference, depth limits). The range is delimited even further by expert-review polygons and compared with that of AquaMaps [@Kaschner:2016tl], OBIS and GBIF occurrence data. The model then determines a species habitat preference based on the assumptions that the relative abundance of a species is determine by the number of habitats in a grid cell and the distance of the species to each habitat, and the importance of the habitat to the species size. Finally, the species equatorial submergence (e.g. the latitudinal region where a species is not seen in between poles) is estimated for each species. See [@Close:2006ux; @Palomares:wg] for model's details.

@@

data.frame(Dasets_names,
           Brieff_Description,
           Sources) %>% 
  kable("latex", booktabs = T,
        caption = "Summary table of data sources used in the current analysis") %>%
  kable_styling(full_width = F) %>%
  column_spec(2,width = "8cm") %>% #wraps text in column2
  column_spec(3,width = "3cm") #%>% #wraps text in column2

@@

### Criteria 1; Neighboring EEZs

As previously mentioned, we define transboundary species as those marine species that happen within the EEZs of two or more neighboring countries. Hence, the first criteria was that the species would only be transboundary between two neighboring countries, regardless of the species extended distribution. This way, a species such as Atlantic cod (*Gadus morhua*) distributed along the north Atlantic, was only considered transboundary between each of the neighboring nations covering its distribution, rather than the region as a whole. We assume that this would reduce the differences between species and stock as is more likely that neighboring species belong to the same stock, rather than, lets say, Cod in Norway and Canada. However, the analysis here presented was kept at the species level and stocks within countries were not considered.

We define the EEZ boundaries using the SAU shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) that sub-divides EEZs by regions (e.g. Mexico Pacific and Mexico Atlantic) and determine the intersections between polygons using the *sf* *R* package [@SF:package]. When estimating transboundary species, we filter out those shared by the EEZs sub-regions, and when aggregating by country we only counted the species once, if it appeared in more than one sub-region. Species that were present in EEZs that were non-continental territories neighboring other countries were kept (e.g. Argentina and Falkland Islands), but removed when the non-continental territory belonged to the same nation (e.g. Brazil and Fernando de Noronha).

@@

We assume that a species was only present in a given grid-cell if it was reported in the SAU catch database. Therefore, all species that were not reported as caught in any single year between the reference years (1970 to 2000) in a given grid-cell were dropped, regardless of the *Species Index*. The assumption relies in that if a commercial species is projected within any fishing country, such species would have been fished, and thus reported at some period of time, thus validating the models.

@@

We estimated the economic contribution of transboundary species for each country using global ex-vessel price [@Tai2017]. The data includes ex-vessel price from multiple sources and a structured interpolation method (e.g. similar countries, species) to fill in data gaps [@Sumaila:2015uc]. The contribution of transboundary species was done by aggregating the catch of each species that happened within the EEZ of the countries sharing that species. For this study, we did not include catch from areas beyong national jurisdiction. Finally, we use the catch data to determine the exploitaiton category of each sepcies within a EEZ. Although this method has previouslly used to estimate stock status [@Grainger:1996tk; @Kleisner:2011wn], the categories presented here are intended to be seen as catch trends, and not status of each species. We only assessed species within each EEZ that had at least ten years of data between the first and last reported landings with at least five consecutive data years. Three categories were drown from the method depending on the year's catch, the maximum catch over time and the minimum catch after the catch peak (**Table XX**). Finally, we reported the predominant category over the last ten years of catch data (2004-2014).

@@

### Original version

# Stock_Status <- c("Rebuilding","Developing","Max Exploited","Over Exploited","Collapsed","No Status")

# Rules <- c(
#   "(Year of Catch > Year Post Max. Min. & Post Max Min Catch < (Max Catch*0.10)) & (Catch > (Max Catch*0.10) & Catch < (Max Catch*0.50))", # Reb.
#   "Year of Catch < Year of Max. Catch & Catch <= (Max Catch*0.50) or Year of Max Catch = Last Year of data)", #Dev.
#   "Catch > (Max Catch*0.50)", #Max Exp.
#   "Year of Catch > Year of Max. Catch & (Catch > (Max. Catch*0.10) & Catch < (Max. Catch*0.50)", # Over.
#   "Year or Catch > Year of Max. Catch & Catch < Max. Catch*0.10", # Colla.
#   "Non of the above rules applied"
# )

# data.frame(Stock_Status,
#            Rules) %>% 
#   kable("latex", booktabs = T,
#         caption = "Rules to determine the status of each transboundary species") %>%
#   kable_styling(full_width = F) %>%
#   column_spec(1,width = "3cm") %>% #wraps text in column2
#   column_spec(2,width = "13cm") #%>% #wraps text in column2

### Grouped version

Stock_Status <- c("A - Growing Fishery","B - Stable Fishery"," C - Fishery under 10% of Peak","No Status")

Rules <- c(
  "(Year of Catch > Year Post Max. Min. & Post Max Min Catch < (Max Catch*0.10)) & (Catch > (Max Catch*0.10) & Catch < (Max Catch*0.50)) or Year of Catch < Year of Max. Catch & Catch <= (Max Catch*0.50) or Year of Max Catch = Last Year of data)", #Rebuilding or #Developing
  "Catch > (Max Catch*0.50)", #Max Expploitation
  "Year of Catch > Year of Max. Catch & (Catch > (Max. Catch*0.10) & Catch < (Max. Catch*0.50) or Catch < Max. Catch*0.10", #Over explpoited or collapsed
  "Non of the above rules applied"
)
